Giant Low Surface Brightness Galaxies As Merger Remnants

### 2016-02-04 glsb.tex
lunch break

\subsection{IFU spectra}
We obtained spatially resolved spectroscopy of our target galaxies using the Mitchell Spectrograph (formerly VIRUS-P)\citep{Hill08} on the 2.7m Harlan J. Smith telescope at McDonald Observatory.  Observations were made during dark time in February, March, and April 2010.  Identical instrument setups were used for all observations.  The Mitchell Spectrograph with the VP-2 IFU bundle used here has a square array of 246 optical fibers which sample the 1.9\arcmin$\times$1.9\arcmin\ field-of-view with a 1/3 filling factor, with each fiber having a diameter of 4.23\arcsec.  The spectrograph images the spectrum of each fiber on a 2048$\times$2048 Fairchild Imaging CCD.

@@

\subsection{GLSB Centers}

@@

% based on WMAP7+BAO+H0 mean
%>>> fidcosmo
%{'tau': 0.087, 'omega_n_0': 0.0, 'sigma_8': 0.809, 'h': 0.704, 'N_nu': 0, 'omega_lambda_0': 0.7274, 'n': 0.963, 'Y_He': 0.24, 'z_reion': 10.4, 'omega_b_0': 0.0456, 'omega_k_0': 0.0, 't_0': 13.75, 'omega_M_0': 0.2726}
%xxx-clearly need to do a galfit to the central objects and show the residuals!  hmm, they are not well fit by simple Sersic profiles

@@

The central regions of GLSBs are often called ``bulges'', however, looking at Figure~\ref{elliptical_relations}, it is clear that these are much brighter than bulges of typical spiral galaxies and the centers of GLSBs are on the bright end of the red-sequence.

@@

We follow \citet{Bernardi03,Bernardi03b} in fitting fundamental plane parameters to our selected SDSS galaxies.  We fit the fundamental plane by minimizing $\chi^2$ in, 
\begin{equation}
\chi^2 = \log R_0 - a \log \sigma - b \log I_0 - c
\end{equation}
where $R_0$ is the effective radius, $\sigma$ is velocity dispersion in kilometers per second and $I_0$ is the surface brightness.  The resulting best-fit parameters are  a =1.376 , b = -0.661, c = -9.938.  We note that we have not made a rigorous volume-complete selection of galaxies, rather, simply galaxies with similar magnitudes, colors, and redshifts to our GLSBs.   See \citet{Saulder13} for a more rigorous analysis of the fundamental plane in SDSS.

\subsection{GLSB Disks}

The diameters of GLSB disks are very large, with Malin~1's stellar disk detected out to a radius of 115 kpc \citep{Moore06} and 80 kpc \citep{Galaz15}.  Stellar disks are typically embedded in much larger \HI\ disks.  The size of the GLSB stellar disks is similar to the largest \HI\ disks observed in nearby galaxies, in support of the \citet{Mapelli08} model where the GLSB disks are the faded remnants of a normal \HI\ disk that has experienced a merger induced star-burst.

For comparison:  \citet{Williams13} trace star formation out to 11 disk scale lengths (radius of 22 kpc) in NGC 2403.

@@

%/astro/store/shared-scratch1/yoachim/Research/GLSB/THINGS/thingsplot.py
\begin{figure*}
\epsscale{.5}
\plotone{Plots/things.eps}
\epsscale{1}
\caption{ Comparison of the GLSB disks with other galaxy disks from the literature.  THINGS data taken from \citet{DeBlok2008}, Malin 1 from \citet{Galaz15}, Malin 2 and UGC 6614 from \citet{Pickering97}.  Here, $R_{max}$ is the maximum radius with an \HI\ detection for the THINGS galaxies, Malin~2, and UGC~6614 and the maximum radius with a stellar flux detection for Malin~1 XXX.  It is interesting to note that the \emph{stellar} disks of GLSB galaxies can be larger than the \HI\ disks of nearby systems.  \label{disk_comp}}
\end{figure*}

\section{Kinematics}

@@

It might be nice to spin this as a chance to measure the DM halo of ellipticals to very large radii.  Some relevant papers for comparison:
\citet{Weij09} use SAURON to get way out.  \citet{Proctor09} use Keck.  \citet{Murphy2011} use VP.

%$R_e = 7.3$\arcsec for UGC06614\\
%$R_e = 7.17$\arcsec for Malin2

The kinematics of the extended GLSB disks represents a unique probe for measuring the extended dark matter halo of elliptical galaxies.

@@

Notes on individual rotation curves:
The Malin 2 rotation curve is rather complicated, having several abrupt rises.  This is especially strange given the very smooth photometric profile.  UGC 6614, on the other hand, has a relatively normal spiral galaxy-like rotation curve that rises linearly out to 10 kpc and then flattens.

The lack of rotation in the central regions of the galaxies is not particularly surprising, as this is the region that is flux-dominated by the central elliptical galaxy.  XXX-that hopefully we have data that shows they are not rotating or rotating in a different orientation.

@@

%using /astro/store/shared-scratch1/yoachim/Research/GLSB/Analysis/mass_estm.pro
If we use the circular speed of the \HI\ data, we end up with masses of $25\times 10^{11} \msun$ for Malin 2 and $6.5\times 10^{11} \msun$ for UGC 6614.  So that's dynamical to stellar mass ratios of 18 and 14 respectively.  Looks like they both have M$_{\rm{HI}}$ around 3-4 $10^{10} \msun$.

Thus, for Malin~2, we reach a radius of $\sim 6.2 R_e$, and find a dark matter fraction of 81\%.  For UGC 6614, we reach a radius of  $\sim 5.2 R_e$ and find a dark matter fraction of 71\%.

\citet{Murphy2011} reach $\sim2 R_e$ using just stellar data, and then rely on globular cluster kinematics to extended out to $5 R_e$.

@@

XXX-It would be great to get high resolution \HI\ observations of these systems since they could probably be higher spatial resolution.

>>>>

We obtained spatially resolved spectroscopy of our target galaxies using the Mitchell Spectrograph (formerly VIRUS-P)\citep{Hill08} on the 2.7m Harlan J. Smith telescope at McDonald Observatory.  Observations were made during dark time in February, March, and April 2010.  Identical instrument setups were used for all observations.  The Mitchell Spectrograph with the VP-2 IFU bundle used here has a square array of 246 optical fibers which sample the 1.9\arcmin$\times$1.9\arcmin\ field-of-view with a 1/3 filling factor, with each fiber having a diameter of 4.23\arcsec.  The spectrograph images the spectrum of each fiber on a 2048$\times$2048 Fairchild Imaging CCD.

@@

\subsection{GLSB Centers}\label{sec:centers}

@@

% based on WMAP7+BAO+H0 mean
%>>> fidcosmo
%{'tau': 0.087, 'omega_n_0': 0.0, 'sigma_8': 0.809, 'h': 0.704, 'N_nu': 0, 'omega_lambda_0': 0.7274, 'n': 0.963, 'Y_He': 0.24, 'z_reion': 10.4, 'omega_b_0': 0.0456, 'omega_k_0': 0.0, 't_0': 13.75, 'omega_M_0': 0.2726}

@@

The central regions of GLSBs are often refered to in the literature as ``bulges'', however, looking at Figure~\ref{elliptical_relations}, it is clear that these are much brighter than bulges of typical spiral galaxies and the centers of GLSBs are on the bright end of the red-sequence.

@@

We follow \citet{Bernardi03,Bernardi03b} in fitting fundamental plane parameters to our selected SDSS galaxies.  We fit the fundamental plane by minimizing $\chi^2$ in, 
\begin{equation}
\chi^2 = \log R_0 - a \log \sigma - b \log I_0 - c
\end{equation}
where $R_0$ is the galaxy's effective radius, $\sigma$ is velocity dispersion in kilometers per second and $I_0$ is the surface brightness.  The resulting best-fit parameters are  a =1.376 , b = -0.661, c = -9.938.  We note that we have not made a rigorous volume-complete selection of galaxies, rather, simply galaxies with similar magnitudes, colors, and redshifts to our GLSBs.   See \citet{Saulder13} for a more rigorous analysis of the fundamental plane in SDSS.

\subsection{GLSB Disks}\label{sec:disks}

The diameters of GLSB disks are very large, with Malin~1's stellar disk detected out to a radius of 80 kpc \citep{Galaz15}.  Stellar disks are typically embedded in much larger \HI\ disks.  The size of the GLSB stellar disks is similar to the largest \HI\ disks observed in nearby galaxies, in support of the \citet{Mapelli08} model where the GLSB disks are the faded remnants of a normal \HI\ disk that has experienced a merger induced star-burst.

%For comparison, \citet{Williams13} trace star formation out to 11 disk scale lengths (radius of 22 kpc) in NGC 2403.

@@

%/astro/store/shared-scratch1/yoachim/Research/GLSB/THINGS/thingsplot.py
\begin{figure*}
\epsscale{.5}
\plotone{Plots/things.eps}
\epsscale{1}
\caption{ Comparison of the GLSB disks with other galaxy disks from the literature.  THINGS data taken from \citet{DeBlok2008}, Malin 1 from \citet{Galaz15}, Malin 2 and UGC 6614 from \citet{Pickering97}.  Here, $R_{max}$ is the maximum radius with an \HI\ detection for the THINGS galaxies, Malin~2, and UGC~6614 and the maximum radius with a stellar flux detection for Malin~1 XXX.  These are all maximum detection radii, and thus should be considered lower limits.  XXX--add in the stellar disks for Malin 2 and UGC 6614. \citet{Kasparova14} trace the light out to 60\arcsec on Malin~2, so 56 kpc\label{disk_comp}.  I trace UGC 6614 out to 65\arcsec, so 29 kpc.}
\end{figure*}

\section{Kinematics}\label{sec:kinematics}

@@

As we have shown in \S\ref{sec:centers} and \S\ref{sec:disks}, GLSB galaxies can be well described as large normal ellipticals that have managed to accrete a high angular momentum disk of stars and gas.  Given the total \HI mass in GLSBs, the accreted object is consistent with a much lower-mass gas-rich spiral.  If this is in fact the way GLSBs form, the disk kinematics represent a unique probe of the dark matter halo of elliptical galaxies.

@@

The Malin 2 rotation curve (Figure~\ref{malin_kinem}) is rather complicated, having several abrupt rises.  This is especially strange given the very smooth photometric profile.  UGC 6614 (Figure~\ref{ugc_kinem}), on the other hand, has a relatively normal spiral galaxy-like rotation curve with a linear rise and flat outer region.  One important difference is that the linear rise extends for 10 kpc.

The lack of rotation in the central regions of the galaxies is not particularly surprising, as this is the region that is flux-dominated by the central elliptical galaxy.

@@

Thus, using just the stellar kinematics, for Malin~2, we reach a radius of $\sim 6.2 R_e$, and find a dark matter fraction of 81\%.  For UGC 6614, we reach a radius of  $\sim 5.2 R_e$ and find a dark matter fraction of 71\%.

\citet{Murphy2011} are able to use IFU observations of M87 to reach $\sim2 R_e$ using just stellar data, and then rely on globular cluster kinematics to extended out to $5 R_e$ where they find a dark matter fraction of 85\%.  \citet{Weij09} is similarly able to use IFU observations to reach $\sim 4 R_e$ on two elliptical galaxies while \citet{Proctor09} reaches $\sim 3 R_e$ using multi-object spectroscopy.

%using /astro/store/shared-scratch1/yoachim/Research/GLSB/Analysis/mass_estm.pro
If we use the circular speed of the \HI\ data from the literature, we end up with masses of $25\times 10^{11} \msun$ for Malin 2 and $6.5\times 10^{11} \msun$ for UGC 6614.  So that's dynamical to stellar mass ratios of 18 and 14 respectively.  %Looks like they both have M$_{\rm{HI}}$ around 3-4 $10^{10} \msun$.

%Some relevant papers for comparison:
%\citet{Weij09} use SAURON to get way out.  \citet{Proctor09} use Keck.  \citet{Murphy2011} use VP.
%$R_e = 7.3$\arcsec for UGC06614\\
%$R_e = 7.17$\arcsec for Malin2

@@

%XXX-It would be great to get high resolution \HI\ observations of these systems since they could probably be higher spatial resolution.
